---
title: "How to create clustered bar graphs with Python"
author: "Ruth Ogal"
date: "2025-03-19"
output: 
  html_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--CSS Rules for styling HTML elements-->
<style>
  table {
    border-collapse: collapse;
    width: 50%;
    margin-bottom: 30px; 
    margin-top: 30px;
  }
  th, td {
    border: 1px solid black;
    padding: 8px;
    text-align: left;
  }
  caption {
    font-weight: bold; /* Make the caption text bold */
    font-style: italic; /* Make the caption text italic */
  }
</style>


```{r, comment=NA, warning=FALSE}
# Load package for calling Python from R
library(reticulate)
```


A clustered bar graph, also known as a grouped bar chart, is used to plot numerical values for the levels of two categorical variables in a dataset. In these graphs, the bars are grouped based on the levels of one categorical variable. The colors of the bars within each group represent the levels of the second categorical variable. In this article, we use a dummy dataset to create clustered bar graphs with Python.

## Pre-requisites

To follow through with the exercises in this article, you should

+ Have a working knowledge of Python.

+ Have the following Python libraries installed in your machine: *pandas*, *matplotlib*, *plotly*, and *seaborn*.


## Import Libraries

Import all libraries necessary to complete this exercise using the following code:

```{python}
# Import libraries
import pandas as pd
import numpy as np
import plotly.express as px
import plotly.graph_objects as go
import matplotlib.pyplot as plt
import seaborn as sns
```


## The Data

<!--Custom HTML Table-->
<table>
  <caption style="caption-side: bottom; text-indent: 30px;">
    Table 1. Dummy sales data.
  </caption>
  <thead>
    <tr>
      <th>Store</th>
      <th>Product</th>
      <th>Sales ($)</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>A</td>
      <td>P1</td>
      <td>42,500</td>
    </tr>
    <tr>
      <td>B</td>
      <td>P1</td>
      <td>48,502</td>
    </tr>
    <tr>
      <td>C</td>
      <td>P1</td>
      <td>39,500</td>
    </tr>
    <tr>
      <td>A</td>
      <td>P2</td>
      <td>43,600</td>
    </tr>
    <tr>
      <td>B</td>
      <td>P2</td>
      <td>32,303</td>
    </tr>
    <tr>
      <td>C</td>
      <td>P2</td>
      <td>29,500</td>
    </tr>
    <tr>
      <td>A</td>
      <td>P3</td>
      <td>49,270</td>
    </tr>
    <tr>
      <td>B</td>
      <td>P3</td>
      <td>49,664</td>
    </tr>
    <tr>
      <td>C</td>
      <td>P3</td>
      <td>48,507</td>
    </tr>
  </tbody>
</table>


Table 1 shows a dummy dataset having 3 columns and 9 rows. The dataset is said to have a long format because its first column (Store) has values repeating in multiple rows. We would like to see how the sales values for products P1, P2, and P3 change across stores A, B, and C, and a clustered bar graph is ideal for visualizing these changes. 
We use pandas to recreate the dataset in Python. Here is the code:


```{python, comment=NA}
# Create dataset
df = pd.DataFrame({"Store": ["A", "B", "C", "A", "B", "C", "A", "B", "C"], "Product": ["P1", "P1", "P1", "P2", "P2", "P2", "P3", "P3", "P3"], "Sales": [42500, 48502, 39500, 43600, 32303, 29500, 49270, 49664, 48507]})
print(df)
```


## Creating the graph

### Method 1: Using Plotly

You can use either the *plotly.express* or the *plotly.graph_objects* module to plot graphs in Plotly. The *plotly.graph_objects* module is best for building complex graphs requiring significant customization.

To build the clustered bar graph in *plotly.express*, use the following code:


```{python, jupyter_compat=TRUE}
# plotly.express clustered barchart
fig = px.bar(df, x="Store", y="Sales",
             color="Product", barmode="group", title="Product sales in $",
             color_discrete_sequence=px.colors.qualitative.Set2,
             width=500, height=400,
             text_auto=True, template="plotly_white")
fig.update_yaxes(visible=False)
fig.show()
```
<p style="font-weight: bold; font-style: italic; text-indent: 30px;">Figure 1. Clustered bar graph created using plotly.express.
</p> 


You can learn more about the parameters of *px.bar()* by using the command *help(px.bar)*. 

In the code above, we use the *update_yaxes()* figure method to suppress the graph’s y-axis and horizontal grid lines, removing graph clutter. The result is displayed in Figure 1. 

Next, we create another clustered bar graph using *plotly.graph_objects*.* However, to use this module, you must reorganize the dataset into a wide format, ensuring that each row in the new dataset represents a unique store. 

As the data is small, you can recreate it from scratch using the following code:

```{python, comment=NA}
df_wide = pd.DataFrame({"Store": ["A", "B", "C"], "P1": [42500, 48502, 39500], 
              "P2": [43600, 32303, 29500], "P3": [49270, 49664, 48507]})
print(df_wide)
```


If the data were large, it’d be easier to reorganize it using the *pivot()* method in *pandas*. This method has the following three crucial parameters:

+ columns: columns of the new DataFrame.

+ index: column used to create the new DataFrame’s index.

+ values: columns used to populate the new DataFrame. 

Use the command *help(df.pivot)* to learn more about *pivot()*.

For this case, you can use *pivot()* to reorganize the data as follows:

```{python, comment=NA}
df_wideP = df.pivot(columns=["Product"], index="Store", values="Sales")
print(df_wideP)
```

Having reorganized the data, we can now use *plotly.graph_objects* to visualize it. Here’s the code. 

```{python, jupyter_compat=TRUE}
data=[go.Bar(name="P1", x=df_wide["Store"], y=df_wide["P1"], text=df_wide["P1"], marker_color="rgb(102, 194, 165)"),
    go.Bar(name="P2", x=df_wide["Store"], y=df_wide["P2"], text=df_wide["P2"], marker_color="rgb(252, 141, 98)"),
    go.Bar(name="P3", x=df_wide["Store"], y=df_wide["P3"], text=df_wide["P3"], marker_color="rgb(141, 160, 203)")]
layout = go.Layout(
    title={"text": "Product sales in $"},
    xaxis={"title": "Store"},
    legend={"title": "Product"},
    width=500, height=400, template="plotly_white",
    barmode="group"
)
fig_go = go.Figure(data=data, layout=layout)
fig_go.update_yaxes(visible=False)
fig_go.show()
```
<p style="font-weight: bold; font-style: italic; text-indent: 30px;">Figure 2: Clustered bar graph created using plotly.graph_objects. </p> 


The code outputs the bar graph in Figure 2.


### Method 2: Using Matplotlib

As we already have a wide *pandas* DataFrame, i.e., df_wide, we can call *pandas’ plot()* method on this DataFrame to plot the required graph. This method leverages *Matplotlib’s* plotting capabilities. Here’s the code:

```{python}
# Clustered bar graph using pandas' plot() method
colors = ["#816e94", "#22a4bf", "#8cc781"]
df_wide.plot(x="Store", 
        kind="bar", width=0.9,
        stacked=False, figsize=(5, 4), title="Product sales in $",
        xlabel="Store", ylabel="Sales", color=colors)
plt.legend(loc="lower right", title="Product");
```
<p style="font-weight: bold; font-style: italic; text-indent: 30px;">Figure 3: Clustered bar graph created using pandas’ plot() method.</p> 


Alternatively, you can use matplotlib’s bar() method to create the graph. Here’s the code:

```{python}
# Clustered bar graph using matplotlib's bar() method
barWidth = 0.25
colors = ["#816e94", "#22a4bf", "#8cc781"]

# Sales values for products p1, p2, and p3
p1 = df_wide["P1"] 
p2 = df_wide["P2"]
p3 = df_wide["P3"]

# Positions of bar groups along the x-axis
pos1 = np.arange(len(p1))  
pos2 = pos1 + barWidth
pos3 = pos2 + barWidth  

# Plot graph
fig, ax = plt.subplots(figsize=(5, 4))
ax.bar(pos1, p1, color=colors[0], width=barWidth, label="P1")
ax.bar(pos2, p2, color=colors[1], width=barWidth, label="P2")
ax.bar(pos3, p3, color=colors[2], width=barWidth, label="P3")
ax.set_xlabel("Store")
ax.set_xticks(pos1 + barWidth)  # Set x-ticks to be in the middle of the bar groups
ax.set_xticklabels(["A", "B", "C"])  # store group labels
ax.legend(loc="lower right", title="Product")
plt.title("Product sales in $")
plt.ylabel("Sales")
plt.show()
```
<p style="font-weight: bold; font-style: italic; text-indent: 30px;">Figure 4: Clustered bar graph created using matplotlib’s bar() method.</p> 


As you can see, this method builds every bar from the ground up. To use it effectively, you must specify the positions of the bar groups and ticks along the x-axis and the width of the bars. In this case, the first, second, and third groups start at 0, 1, and 2, respectively. Each group contains three bars (for P1, P2, and P3), so the total width of these three bars must be less than 1 unit, with each bar having a maximum width of 0.33 units. If the width of a bar exceeds 0.33 units, the bar groups will touch each other or overlap. We set the bar width at 0.25 units, resulting in the graph shown in Figure 4. 


### Method 3: Using Seaborn

The *Seaborn* library is built on top of *Matplotlib* and provides many tools to enhance *Matplotlib’s* plotting capabilities.

We use *Seaborn’s barplot()* method to plot the clustered bar graph. This method supports both wide- and long-format data. However, it’s easier to use with long-format data where critical variables are not spread over multiple columns. Here’s the code:

```{python}
# Clustered bar graph using Seaborn's barplot() method
custom_palette = ["#a6cee3", "#1f78b4", "#b2df8a"]
fig, ax = plt.subplots(figsize=(5, 4)) # Set plot dimensions.
sns.set_style("white")
sns.barplot(x="Store", y="Sales", hue="Product", data=df, palette=custom_palette).set(title="Product sales in $")
plt.legend(loc="lower right")
plt.show()
```
<p style="font-weight: bold; font-style: italic; text-indent: 30px;">Figure 5: Clustered bar graph created using Seaborn’s barplot() method.</p> 


Many methods used in the code above, like *plt.show()* and *plt.subplots()*, are familiar, as we’ve already used them when plotting with *Matplotlib*.

The code outputs the graph shown in Figure 5.

## Conclusion

Being Python’s first visualization library, *Matplotlib* has become a favorite of many Python users. 
Several later libraries, like *pandas* and *Seaborn*, are built on top of it because it is highly customizable and flexible. However you may have to write a lot of code with *Matplotlib* to create appealing graphs. Generally, of the three libraries discussed in this article, *Plotly* is the easiest to use for creating clustered bar graphs and provides the most appealing results. 


